Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
more

\newcommand{\absconstraints}{X}

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, %we claim that
$\conv(\varphi)$ is equisatisfiable with $\varphi$ for any input $\varphi$.
%\ct{Must define "equisatisfiable".
%What we mean here is a weak form of satisfiability, we could instead say that
%$\conv(\varphi)$ is $T$-satisfiable iff $\varphi$ is $T$-satisfiable.
%} 
%\jb{We do define it, even it is weak form.}
The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable or unsatisfiable response from an SMT solver on input $\conv(
\varphi )$ implies the existence or nonexistence of a model for
$\varphi$. Moreover, the models of $\conv( \varphi )$ contain
pertinent information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) described above interprets $\con{c}$ as $14$
and $\con{s}(n) = \sum_{i=1}^n i$ for $0 \le n \le 14$,
and there exists a model of formula~(\ref{eq:ex-before}) that also interprets
$\con{c}$ and and $\con{s}(n)$ in the same way (for $0 \le n \le 14$).
In general, there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma} - \sfundefs{\Sigma}$, i.e., $\{
\con{s} \}$ in the example. Furthermore, the model of $\conv( \varphi )$ will
also give correct information for the defined functions at all points belonging
to the domains of the corresponding abstract types. This can sometimes help
users debug their specifications.

@@

In the following, for a ground formula $\varphi$, 
we write $\absconstraints( \varphi )$ to denote the set of constraints 
that force our concretization functions to have the necessary elements in their range
for determining the satisfiability of $\varphi$ with respect to function definitions in our translation.
Formally, we define $\absconstraints( \varphi )$ as:
\[
\absconstraints( \varphi ) = \{ \exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \mid \con{f}( \vec t ) \in \terms( \varphi ), \con{f} \in \sfundefs{\Sigma} \}
\]
The following lemma states the central invariant behind our translation $\conv$.

@@

\begin{cor}\label{cor:conv}
If $\varphi$ is a formula not containing function definitions, then $\conv( \varphi )$ entails $\varphi$.
\end{cor}

\begin{theorem}\label{thm:equi}
If $\varphi$ is a\/ $\Sigma$-formula that is in definitional form with respect to\/
$\sfundefs{\Sigma}$
and the set of function definitions\/ $\DDD$ corresponding to functions in
$\sfundefs{\Sigma}$ is admissible, then\/ $\varphi$ and\/ $\conv( \varphi )$ are
equisatisfiable.
\end{theorem}
\begin{proofsketch}
First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \stypes{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \cdots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$,
interprets $\fargtype{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \cdots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Since $\J$ satisfies $\varphi$, it satisfies a set of ground literals $L$ that entail $\varphi$.
Furthermore, $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
%By case analysis it follows that $\conv_0$ maintains the invariant that
%if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( {t'}, \chi )$,
%then $\I$ satisfies $t$ if and only if $\J$ satisfies $t'$.
Thus, $\J$ satisfies $L \cup \absconstraints( L )$, and by Lemma~\ref{lem:conv} we conclude $\J$ satisfies $\conv( \varphi )$.

>>>>

\newcommand{\absconstraints}{\mathrm{X}}

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, %we claim that
$\conv(\varphi)$ is equisatisfiable with $\varphi$ for any input $\varphi$.
%\ct{Must define "equisatisfiable".
%What we mean here is a weak form of satisfiability, we could instead say that
%$\conv(\varphi)$ is $T$-satisfiable iff $\varphi$ is $T$-satisfiable.
%} 
%\jb{We do define it, even it is weak form.}
The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable or unsatisfiable response from an SMT solver on input $\conv(
\varphi )$ implies the existence or nonexistence of a model for
$\varphi$. Moreover, the models of $\conv( \varphi )$ contain
pertinent information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) given above interprets $\con{c}$ as $14$
and $\con{s}(n) = \sum_{i=1}^n i$ for $0 \le n \le 14$,
and there exists a model of formula~(\ref{eq:ex-before}) that also interprets
$\con{c}$ and and $\con{s}(n)$ in the same way (for $0 \le n \le 14$).
In general, there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma} - \sfundefs{\Sigma}$, i.e., $\{
\con{s} \}$ in the example. Furthermore, the model of $\conv( \varphi )$ will
also give correct information for the defined functions at all points belonging
to the domains of the corresponding abstract types. This can sometimes help
users debug their specifications.

@@

In the following, for a set of ground literals $L$, 
we write $\absconstraints( L )$ to denote the set of constraints that
force the concretization functions to have the necessary elements in their
range for determining the satisfiability of $L$ with respect to function
definitions in the translation.
Formally, we define
$\absconstraints( L )$
as
$\{ \exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t
\mid \con{f}( \vec t ) \in \terms( L ),\: \con{f} \in \sfundefs{\Sigma}
\}$.
The following lemma states the central invariant behind the translation $\conv$.

@@

\begin{cor}\label{cor:conv}
If $\varphi$ is a formula not containing function definitions, then $\extendsig{\Sigma} = \Sigma$
and $\conv( \varphi )$ entails $\varphi$.
\end{cor}

\begin{theorem}\label{thm:equi}
If $\varphi$ is a\/ $\Sigma$-formula that is in definitional form with respect to\/
$\sfundefs{\Sigma}$
and the set of function definitions\/ $\DDD$ corresponding to functions in\/
$\sfundefs{\Sigma}$ is admissible, then\/ $\varphi$ and\/ $\conv( \varphi )$ are
equisatisfiable.
\end{theorem}
\begin{proofsketch}
First, we show that if $\varphi$ is satisfied by $\Sigma$-interpretation $\I$, then $\conv( \varphi )$ is satisfied by a $\extendsig{\Sigma}$-interpretation $\J$.
Let $\J$ be the $\extendsig{\Sigma}$-interpretation that interprets all types $\tau \in \stypes{\Sigma}$ as $\tau^\I$,
all functions $\con{f} \in \sfuns{\Sigma}$ as $\con{f}^\I$,
and for each function $\con{f}$ of type $\tau_1 \times \cdots \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$,
interprets $\fargtype{\con{f}}$ as a tuple of values of type $\tau_1^\I \times \cdots \times \tau_n^\I$,
and each $\fargx{\con{f}}{i}$ as the $i^{th}$ projection on such tuples for each $i = 1, \ldots, n$.
Since $\J$ satisfies $\varphi$, it satisfies a set of ground literals $L$ that entail $\varphi$.
Furthermore, $\J$ satisfies \emph{every} constraint of the form $\exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$,
since by our construction of $\J$ there exists an $a$ such that $a^\J = \vec t^\J$.
%By case analysis it follows that $\conv_0$ maintains the invariant that
%if $\conv_0( t\typ{\Bool}, \ppos )$ returns $( {t'}, \chi )$,
%then $\I$ satisfies $t$ if and only if $\J$ satisfies $t'$.
Thus, $\J$ satisfies $L \cup \absconstraints( L )$, and by Lemma~\ref{lem:conv} we conclude $\J$ satisfies $\conv( \varphi )$.
